Shifting Shadows of the Six: Decoding Toronto's Crime Trends from 2014 to 2023

### 2024-01-24 outputs/paper/paper.qmd
Update paper content, especially for data section and discussion section

---
title: "Shifting Shadows of the Six: Decoding Toronto's Crime Trends from 2014 to 2023"
author: 
  - Yuanyi (Leo) Liu
thanks: "Code and data are available at: https://github.com/leoyliu/STA302-Paper-1."
date: today
date-format: long
abstract: "In this study, we conducted a comprehensive analysis of crime trends in Toronto, utilizing a dataset detailing various criminal activities from 2014 to 2023. Our investigation focused on the evolution of different types of crimes, with a specific emphasis on contrasting violent crimes, such as assault and robbery, against non-violent offenses like theft and fraud. The findings reveal significant patterns in both categories, highlighting an evolving landscape of criminal behavior in Toronto. This research is crucial as it sheds light on the changing dynamics of urban crime, providing valuable insights for policymakers, law enforcement agencies, and community leaders in developing targeted strategies for crime prevention and enhancing public safety in metropolitan areas."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

# install.packages("knitr")
# install.packages("janitor")
# install.packages("lubridate")
# install.packages("opendatatoronto")
# install.packages("tidyverse")
# install.packages("palmerpenguins")
# install.packages("here")

library(knitr)
library(janitor)
library(lubridate)
library(opendatatoronto)
library(tidyverse)
library(palmerpenguins)
library(here)
```

@@

The paper is structured to facilitate a comprehensive understanding of the study and its implications. Following this introduction @sec-intro, the second section presents the data @sec-data, detailing the data sources, analytical techniques, and the rationale behind the chosen methods. The third section discusses the results @sec-result, elaborating on the observed trends and patterns in crime data. The fourth section provides an in-depth discussion @sec-discussion of these findings, exploring potential factors influencing these trends and drawing connections to broader urban issues. Finally, the paper concludes with a summary of the findings, their implications for urban safety and policy, and suggestions for future research in this area.

@@

## Source of the Data
The dataset comprises crime statistics in Toronto, spanning from 2014 to 2023. It was sourced from Open Data Toronto[@cite1], a reputable provider of public safety and criminal statistics. This dataset is publicly available and is often used for academic and policy research due to its reliability and comprehensive coverage.

@@

## Data Manipulation and Cleaning in R [@citeR]
The data was processed and cleaned using R, a powerful statistical programming language. Key operations included:

* **Data Cleaning:** Addressing missing values, standardizing formats, and filtering out irrelevant variables (e.g., removing rate-based columns).

@@

@@

TODO: include a summary chart later.

@@

@@

## Data Visualization
For visual representation, `ggplot2`, a versatile package in R for data visualization, was employed. It was used to create comprehensive line plots depicting the trends of various crime categories over the years. These plots are instrumental in visualizing the temporal patterns and changes in crime statistics.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false
```

@@

* **Variation in Crime Trends:** The dataset reveals distinct trends for different types of crimes in Toronto. Some crime categories exhibit an upward trend, indicating an increase in incidents over the years, while others show a downward trend, suggesting a decrease in occurrences. This variation highlights the complexity of crime dynamics in an urban setting.

* **Potential Factors Influencing Trends:** The observed trends in crime rates could be influenced by a multitude of factors. These include socio-economic changes, urban development, law enforcement strategies, policy reforms, and technological advancements in crime reporting. For example, an increase in certain types of crimes might be linked to economic downturns, demographic shifts, or changes in urban density, while a decrease could reflect effective policing strategies or community intervention programs.

## Implications for Policymakers and Law Enforcement: {#sec-first-point}

@@

## Second discussion point

>>>>

---
title: "Shifting Shadows of the Six: Decoding Toronto's Crime Trends from 2014 to 2023"
author: 
  - Yuanyi (Leo) Liu
thanks: "Code and data are available at: https://github.com/leoyliu/STA302-Paper-1."
date: today
date-format: long
abstract: "In this study, we conducted a comprehensive analysis of crime trends in Toronto, utilizing a dataset detailing various criminal activities from 2014 to 2023. Our investigation focused on the evolution of different types of crimes, with a specific emphasis on contrasting violent crimes, such as assault and robbery, against non-violent offenses like theft. The findings reveal significant patterns in both categories, highlighting an evolving landscape of criminal behavior in Toronto. This research is crucial as it sheds light on the changing dynamics of urban crime, providing valuable insights for policymakers, law enforcement agencies, and community leaders in developing targeted strategies for crime prevention and enhancing public safety in metropolitan areas."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

# install.packages("knitr")
# install.packages("opendatatoronto")
# install.packages("tidyverse")
# install.packages("here")
# install.packages("tidyr")
# install.packages("dplyr")
# install.packages("kableExtra")

library(knitr)
library(opendatatoronto)
library(tidyverse)
library(here)
library(tidyr)
library(dplyr)
library(kableExtra)
library(janitor)
```

@@

The paper is structured to facilitate a comprehensive understanding of the study and its implications. Following this introduction @sec-intro, the second section presents the data @sec-data, detailing the data sources, analytical techniques, and the rationale behind the chosen methods. The third section discusses the results @sec-result, elaborating on the observed trends and patterns in crime data. The fourth section provides an in-depth discussion @sec-discussion of these findings, exploring potential factors influencing these trends, drawing connections to broader urban issues, and suggestions for future research in this area.

@@

## Source of the Data
The dataset comprises crime statistics in Toronto, spanning from 2014 to 2023. The dataset analyzed in this report was obtained in csv format from the City of Toronto Open Data Portal[@citedata] using the R package `opendatatoronto`[@citeopendatapkg]. The dataset was last updated on January 14th, 2024. This dataset is publicly available and is often used for academic and policy research due to its reliability and comprehensive coverage.

@@

## Data Manipulation and Cleaning in R [@citeR]
The data was processed and cleaned using R[@citeR], a powerful statistical programming language. Key operations included:

* **Data Cleaning:** Addressing missing values, standardizing formats, and filtering out irrelevant variables using the `tidyverse`[@citetidyverse] package(e.g., removing rate-based columns and NAs).

@@

* **Data Visualization:** For visual representation, `ggplot2` [@citeggplot], a versatile package in R for data visualization, was employed. It was used to create comprehensive line plots depicting the trends of various crime categories over the years.

* **Paper Generation:** The packages `knitr` [@citeknitr] is used to generate the R markdown report.

@@

```{r}
#| label: tabledatasummary
#| echo: false

@@

# Creating the summary table
summary_table <- crime_data_summary %>%
  group_by(Crime_Type, Year) %>%
  summarise(Total_Cases = sum(Total, na.rm = TRUE), .groups = "drop") %>%
  spread(key = Year, value = Total_Cases)

# Define a mapping for crime types
readable_crime_types <- c('ASSAULT' = 'Assault',
                          'AUTOTHEFT' = 'Auto Theft',
                          'BIKETHEFT' = 'Bike Theft',
                          'BREAKENTER' = 'Break and Enter',
                          'HOMICIDE' = 'Homicide',
                          'ROBBERY' = 'Robbery',
                          'SHOOTING' = 'Shooting Incidents',
                          'THEFTFROMMV' = 'Theft from Motor Vehicles',
                          'THEFTOVER' = 'Major Theft')

# Update the Crime_Type column
summary_table$Crime_Type <- readable_crime_types[summary_table$Crime_Type]

# Make a beautiful table
summary_table %>%
  slice(1:10) %>%
  kable(
    caption = "Crime Incidents Summary by Year",
    col.names = c("Crime Type", "2014", "2015", "2016", "2017", "2018", "2019", "2020", "2021", "2022", "2023"),
    digits = 1,
    booktabs = TRUE,
    linesep = "",
    align = c("l", "l", "c", "c", "r", "r"),
    format.args = list(big.mark = ",")) %>%
    kable_styling(latex_options = "HOLD_position", font_size = 10, position = "center")
```
Table 1 organizes data into a clear format where each row indicates a specific type of crime. The columns are segmented by year, spanning from 2014 to 2023. Each cell within the table provides a count of the incidents reported for that particular crime in the corresponding year.

Then, we want to categorize our data into one of two groups: 'Violent' or 'Non-Violent'. This categorization was based on the nature of the crime. For example, crimes like 'ASSAULT', 'HOMICIDE', etc., were categorized as 'Violent', while others like 'THEFT', were categorized as 'Non-Violent'. After categorization, the dataset was aggregated based on these two new categories. This involved summing up the incident counts for each category (Violent and Non-Violent) for every year.

```{r}
#| label: combineddatasummary
#| echo: false
summary_table$Category <- ifelse(summary_table$Crime_Type %in% c("Assault", "Break and Enter", "Homicide", "Robbery", "Shooting Incidents"), "Violent", "Non-Violent")

# Group data by the new Category and sum the cases for each year
two_types_crime <- summary_table %>%
  group_by(Category) %>%
  summarise(across(`2014`:`2023`, ~sum(.x, na.rm = TRUE)))

# Make a beautiful table
two_types_crime %>%
  slice(1:10) %>%
  kable(
    caption = "Crime Incidents Summary by Year",
    col.names = c("Crime Type", "2014", "2015", "2016", "2017", "2018", "2019", "2020", "2021", "2022", "2023"),
    digits = 1,
    booktabs = TRUE,
    linesep = "",
    align = c("l", "l", "c", "c", "r", "r"),
    format.args = list(big.mark = ",")) %>%
    kable_styling(latex_options = "HOLD_position", font_size = 10, position = "center")
```

Table 2 organized view of crime statistics over a range of years, specifically differentiated into two main categories: 'Violent' and 'Non-Violent' crimes. Each row in the table represents one of these two categories. The columns of the table are divided by year, ranging from 2014 to 2023. Under each year's column, the table displays the total number of reported incidents for each crime category. For example, under the column for 2014, you will find the total number of violent crimes reported in that year followed by the total number of non-violent crimes for the same year.

@@

```{r}
#| label: fig-trends-in-two-types-crime
#| fig-cap: Comparative Trend Analysis of Violent and Non-Violent Crime Cases from 2014 to 2023
#| echo: false

# Ensure the data is ungrouped
two_types_crime <- ungroup(two_types_crime)

# Pivot the data to long format for plotting
# Explicitly select the year columns to avoid issues with hidden grouping columns
long_data <- two_types_crime %>%
  select(Category, `2014`:`2023`) %>%
  pivot_longer(cols = -Category, names_to = "Year", values_to = "Cases")

# Create the plot
crime_plot <- ggplot(long_data, aes(x = Year, y = Cases, group = Category, color = Category)) +
  geom_line(linewidth = 1) + # Use linewidth instead of size
  geom_point(size = 3) + # Point size
  theme_minimal() + # Minimal theme
  labs(title = "Evolution of Violent and Non-Violent Crimes Over the Years",
       x = "Year",
       y = "Number of Cases") +
  scale_color_manual(values = c("Violent" = "red", "Non-Violent" = "blue")) +
  theme(
    plot.title = element_text(hjust = 0.5), # Center the plot title
    legend.title = element_blank(), # Remove the legend title
    axis.text.x = element_text(angle = 45, hjust = 1) # Angle the x-axis text for readability
  )

# Print the plot
print(crime_plot)
```

@fig-trends-in-two-types-crime compares the number of cases reported for violent and non-violent crimes from 2014 to 2023. The x-axis represents the years, and the y-axis indicates the number of cases.

The red line represents violent crimes, and the blue line represents non-violent crimes. The plot shows distinct trends for both categories over the specified time frame.

* **Violent Crimes (Red Line):** There is an overall increasing trend in the number of violent crime cases over the years. There are some fluctuations, with a noticeable dip occurring in one of the years, but the general direction is upwards, especially towards the end of the period, indicating a rise in violent crimes or an increase in reporting.

* **Non-Violent Crimes (Blue Line):** The trend in non-violent crime cases shows some variability but does not have as clear an increasing pattern as violent crimes. There are years where the rate has increased or decreased, but the last year shows a notable increase, suggesting a possible upward trend at the end of the period.

Overall, the graph indicates that while both violent and non-violent crimes have seen rises and falls, the increase in violent crimes towards the later years is more pronounced. This could reflect changes in societal conditions, law enforcement practices, or reporting mechanisms.

@@

## Variation in Crime Trends: {#sec-first-point}

The dataset reveals distinct trends for different types of crimes in Toronto. Some crime categories exhibit an upward trend, indicating an increase in incidents over the years, while others show a downward trend, suggesting a decrease in occurrences. This variation highlights the complexity of crime dynamics in an urban setting.

## Potential Factors Influencing Trends: {#sec-second-point}

The observed trends in crime rates could be influenced by a multitude of factors. These include socio-economic changes, urban development, law enforcement strategies, policy reforms, and technological advancements in crime reporting. For example, an increase in certain types of crimes might be linked to economic downturns, demographic shifts, or changes in urban density, while a decrease could reflect effective policing strategies or community intervention programs.

@@
